---
title: "Lecture 4: tidyverse for data manipulation"
subtitle: "Data: Hollywood movies"
engine: knitr
filters:
  - ../../../../_extensions/coatless/webr/webr.lua
format:
  revealjs:
    theme: default
    slide-number: true
    transition: fade
    toc: false
    self-contained: false
    smaller: false
    scrollable: true
    code-font-size: 1.15em
    chalkboard:
      buttons: true
editor_options:
  chunk_output_type: inline
---

```{r setup}
#| include: false
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)
```


## Prep work

- [Tidyverse](https://www.tidyverse.org/packages/) is a popular collection of R packages for working with data in a **tidy** format: each variable is a column, each observation is a row, and each value is a cell.

```{r}
library(tidyverse)
library(Lock5Data)
```


- We'll use the `HollywoodMovies` dataset.
```{r}
#| eval: false
str(HollywoodMovies)
head(HollywoodMovies)
?HollywoodMovies
```


## Review questions
- How many genres are there?  How many movies are in each genre?
- What are the 10 movies with largest difference (in absolute value) between RottenTomatoes and AudienceScore?
- Compare the mean AudienceScore by Genre of the movies.
- What are the five action movies with the largest `WorldwideBO`?
- Which movie has the largest `WorldwideBO` in each genre?



## A tidyverse workflow

Tidyverse functions take a data frame as their first input and return a data frame, so functions can be chained with the native R pipe `|>` or tidyverse pipe `%>%`.

```{r}
#| eval: false
HollywoodMovies |>
  filter(Genre == "Action") |>
  select(Movie, Genre, WorldwideBO)
```

Read the pipe as **“then”**: start with `HollywoodMovies`, then keep action movies, then select the  three named columns.


## Subsetting: **'['**
The square bracket syntax can be used as a function, with the function name **'['**
```{r}
#| eval: false
HollywoodMovies$Movie[1:5]
'['(HollywoodMovies$Movie, 1:5)  # as a function
HollywoodMovies$Movie |> '['(1:5) # pipe version

HollywoodMovies[5,1]
HollywoodMovies |> '['(5,1)

HollywoodMovies[5,]
HollywoodMovies |> '['(5,)
```




## Subsetting: **filter**
```{r}
#| eval: false
# Get info of action movies
HollywoodMovies |>
  filter(Genre == "Action")

# Multiple conditions
HollywoodMovies |>
  filter(Genre == "Action", Profitability > 500)

HollywoodMovies |>
  filter(Genre == "Action" | Genre == "Adventure")
```

(Use `?dplyr::filter` to read the help page)


## Subsetting: **slice** and **select**

```{r}
#| eval: false
# Keep selected rows by position
HollywoodMovies |>
  slice(c(1, 3, 5))

# Select top rows
HollywoodMovies |>
  slice_head(n = 5)

# Keep or drop selected columns
HollywoodMovies |>
  select(Movie, Genre, WorldwideBO)

HollywoodMovies |>
  select(-Distributor, -TheatersOpen)
```

## Sampling: `slice_sample()`

```{r}
#| eval: false
# Randomly sample 10% of rows, without replacement
HollywoodMovies |>
  slice_sample(prop = 0.1)

# Sample five rows, allowing repeats
HollywoodMovies |>
  slice_sample(n = 5, replace = TRUE)
```


## Sorting: `arrange()` and `slice_max()`

```{r}
#| eval: false
# Smallest five worldwide grosses
HollywoodMovies |>
  arrange(WorldwideBO) |>
  slice_head(n = 5)

# Largest five worldwide grosses
HollywoodMovies |>
  slice_max(order_by = WorldwideBO, n = 5, na_rm = TRUE) |>
  select(Movie, Genre, WorldwideBO)

HollywoodMovies |>
  arrange(desc(WorldwideBO)) |>
  slice(1:5)

# Top three values of a derived quantity; ties are retained by default
HollywoodMovies |>
  mutate(score_difference = abs(RottenTomatoes - AudienceScore)) |>
  slice_max(order_by = score_difference, n = 3)
```


## Summary statistics: summarise()
```{r}
#| eval: false
HollywoodMovies |>
  summarise(mean_worldwide = mean(WorldwideBO, na.rm = TRUE), 
            median_worldwide = median(WorldwideBO, na.rm = TRUE), 
            movie_count = n())
```


## Grouping: group_by()

If you are only interested in action movies, say, you might use `filter()`.
```{r}
#| eval: false
HollywoodMovies |>
  filter(Genre == "Action") |>
  summarise(mean_worldwide = mean(WorldwideBO, na.rm = TRUE), 
            median_worldwide = median(WorldwideBO, na.rm = TRUE), 
            movie_count = n())
```


If you now want to generalize to other genres, rather than manually changing "Action" to other genres, you can group by genre instead.

```{r}
#| eval: false
HollywoodMovies |>
  group_by(Genre) |>
  summarise(mean_worldwide = mean(WorldwideBO, na.rm = TRUE), 
            median_worldwide = median(WorldwideBO, na.rm = TRUE), 
            movie_count = n(),
            .groups = "drop")
```

The `.groups = "drop"` argument ensures that the resulting data frame is no longer grouped by genre, so that later operations are not affected.

Can group by multiple variables.

```{r}
#| eval: false
HollywoodMovies |>
  group_by(Genre, Year) |>
  summarise(mean_worldwide = mean(WorldwideBO, na.rm = TRUE), 
            median_worldwide = median(WorldwideBO, na.rm = TRUE), 
            movie_count = n(),
            .groups = "drop")
```




## Example: top 5 most profitable genres

```{r}
#| code-fold: true
#| code-summary: "Show code"
HollywoodMovies |>
  filter(!is.na(Profitability)) |>
  group_by(Genre) |>
  summarise(
    mean_profit = mean(Profitability),
    movie_count = n(),
    .groups = "drop"
  ) |>
  slice_max(order_by = mean_profit, n = 5)
```

Note: nothing is stored; no change is made in the data frame.


## Modify and create variables: mutate

```{r}
#| eval: false
# Create a derived column and select the five largest values
HollywoodMovies |>
  mutate(score_gap = abs(RottenTomatoes - AudienceScore)) |>
  slice_max(order_by = score_gap, n = 5) |>
  select(Movie, RottenTomatoes, AudienceScore, score_gap)

# Save the updated data frame using explicit assignment
movies_with_gap <- HollywoodMovies |>
  mutate(score_gap = abs(RottenTomatoes - AudienceScore))

# Keep only selected columns while creating a derived column
HollywoodMovies |>
  transmute(Movie, score_gap = abs(RottenTomatoes - AudienceScore))

# Rename existing columns
HollywoodMovies |>
  rename(open_weekend_gross = OpenBO, theaters_at_open = TheatersOpen)
```


## Tibble

A **tibble** is a modern version of an R data frame. It prints a compact preview and preserves familiar data-frame behavior while integrating with tidyverse functions.

```{r}
#| eval: false
tbl_movie <- tibble::as_tibble(HollywoodMovies)
class(tbl_movie)
```


## Review question solutions

### 1. Movie counts by genre

How many genres are there, and how many movies are in each genre?

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Show code"
# Concise approach
HollywoodMovies |>
  count(Genre, name = "movie_count", sort = TRUE)

# Alternative using group_by(), summarise(), and n()
HollywoodMovies |>
  group_by(Genre) |>
  summarise(movie_count = n(), .groups = "drop") |>
  arrange(desc(movie_count))
```

### 2. Largest score differences

What are the 10 movies with the largest absolute difference between `RottenTomatoes` and `AudienceScore`?

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Show code"
# slice_max() approach
HollywoodMovies |>
  mutate(score_gap = abs(RottenTomatoes - AudienceScore)) |>
  slice_max(order_by = score_gap, n = 10) |>
  select(Movie, RottenTomatoes, AudienceScore, score_gap)

# Alternative using arrange() and slice()
HollywoodMovies |>
  mutate(score_gap = abs(RottenTomatoes - AudienceScore)) |>
  arrange(desc(score_gap)) |>
  slice(1:10) |>
  select(Movie, RottenTomatoes, AudienceScore, score_gap)
```

### 3. Mean audience score by genre

Compare the mean `AudienceScore` by genre. This uses the grouped summary method introduced above.

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Show code"
HollywoodMovies |>
  group_by(Genre) |>
  summarise(
    mean_audience = mean(AudienceScore, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(desc(mean_audience))
```

### 4. Highest-grossing action movies

What are the five action movies with the largest `WorldwideBO`?

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Show code"
# slice_max() approach
HollywoodMovies |>
  filter(Genre == "Action") |>
  slice_max(order_by = WorldwideBO, n = 5) |>
  select(Movie, Genre, WorldwideBO)

# Alternative using arrange() and slice()
HollywoodMovies |>
  filter(Genre == "Action") |>
  arrange(desc(WorldwideBO)) |>
  slice(1:5) |>
  select(Movie, Genre, WorldwideBO)
```

### 5. Highest-grossing movie in each genre

Which movie has the largest `WorldwideBO` in each genre? `slice_max()` retains ties.

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Show code"
HollywoodMovies |>
  group_by(Genre) |>
  slice_max(order_by = WorldwideBO, n = 1) |>
  ungroup() |>
  select(Movie, Genre, WorldwideBO)
```







