---
title: "Lecture 4: Hollywood movies"
subtitle: "Tidyverse for data manipulation"
engine: knitr
filters:
  - ../../../../_extensions/coatless/webr/webr.lua
format:
  revealjs:
    theme: default
    slide-number: true
    transition: fade
    toc: false
    self-contained: false
    smaller: true
    scrollable: true
    code-font-size: 1.15em
    chalkboard:
      buttons: true
editor_options:
  chunk_output_type: inline
---

```{r setup}
#| include: false
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)
```


## Prep work

- **tidyverse** packages: https://www.tidyverse.org/packages/

- A popular collection of R packages for working with data in a **tidy** format: each variable is a column, each observation is a row, and each value is a cell.

```{r}
library(tidyverse)
library(Lock5Data)
```


- We'll use the `HollywoodMovies` dataset.
```{r}
#| eval: false
str(HollywoodMovies)
head(HollywoodMovies)
?HollywoodMovies
```

## A tidyverse workflow

Tidyverse functions take a data frame as their first input and return a data frame, so functions can be chained with the native R pipe `|>`.

```{r}
#| eval: false
HollywoodMovies |>
  filter(Genre == "Action") |>
  select(Movie, Genre, WorldwideBO)
```

Read the pipe as **“then”**: start with `HollywoodMovies`, keep action movies, then keep the three named columns.

## Choose rows and columns

- `filter()` keeps rows that meet logical conditions.
- `select()` keeps or drops columns.
- `slice()` chooses rows by position.

```{r}
#| eval: false
HollywoodMovies |>
  filter(Genre == "Comedy", RottenTomatoes >= 80) |>
  select(Movie, Genre, RottenTomatoes)

HollywoodMovies |>
  slice(1:5)
```

## Create and order variables

- `mutate()` creates or changes columns.
- `arrange()` sorts rows; use `desc()` for descending order.
- `slice_max()` selects rows with the largest values of a column.

```{r}
#| eval: false
HollywoodMovies |>
  mutate(score_gap = abs(RottenTomatoes - AudienceScore)) |>
  arrange(desc(score_gap)) |>
  select(Movie, RottenTomatoes, AudienceScore, score_gap) |>
  slice_head(n = 5)
```

## Summarize by group

- `group_by()` defines groups for subsequent calculations.
- `summarise()` returns one row per group.
- `count()` is a concise way to count rows in each group.

```{r}
#| eval: false
HollywoodMovies |>
  group_by(Genre) |>
  summarise(
    mean_audience = mean(AudienceScore, na.rm = TRUE),
    movie_count = n(),
    .groups = "drop"
  ) |>
  arrange(desc(mean_audience))

HollywoodMovies |>
  count(Genre, sort = TRUE)
```


## Review questions
- How many genres are there?  How many movies are in each genre?
- What are the 10 movies with largest difference (in absolute value) between RottenTomatoes and AudienceScore?
- Compare the mean AudienceScore by Genre of the movies.
- What are the five action movies with the largest `WorldwideBO`?
- Which movie has the largest `WorldwideBO` in each genre?



## Subsetting: **'['**
The square bracket syntax can be used as a function, with the function name **'['**
```{r}
#| eval: false
HollywoodMovies$Movie[1:5]
'['(HollywoodMovies$Movie, 1:5)  # as a function
HollywoodMovies$Movie |> '['(1:5) # pipe version

HollywoodMovies[5,1]
HollywoodMovies |> '['(5,1)

HollywoodMovies[5,]
HollywoodMovies |> '['(5,)
```




## Subsetting: **filter**
```{r}
#| eval: false
# Get info of action movies
HollywoodMovies |>
  filter(Genre == "Action")

# Multiple conditions are combined with AND
HollywoodMovies |>
  filter(Genre == "Action", Profitability > 500)

# OR conditions use |
HollywoodMovies |>
  filter(Genre == "Action" | Genre == "Adventure")
```

(Use `?dplyr::filter` to read the help page.)


## Subsetting: **slice** and **select**

```{r}
#| eval: false
# Keep selected rows by position
HollywoodMovies |>
  slice(c(1, 3, 5))

# Keep or drop selected columns
HollywoodMovies |>
  select(Movie, Genre, WorldwideBO)

HollywoodMovies |>
  select(-Distributor, -TheatersOpen)
```

## Sampling: `slice_sample()`

```{r}
#| eval: false
# Randomly sample 10% of rows, without replacement
HollywoodMovies |>
  slice_sample(prop = 0.1)

# Sample five rows, allowing repeats
HollywoodMovies |>
  slice_sample(n = 5, replace = TRUE)
```


## Sorting and selecting top rows

```{r}
#| eval: false
# Smallest five worldwide grosses
HollywoodMovies |>
  arrange(WorldwideBO) |>
  slice_head(n = 5)

# Largest five worldwide grosses
HollywoodMovies |>
  slice_max(order_by = WorldwideBO, n = 5, na_rm = TRUE) |>
  select(Movie, Genre, WorldwideBO)

# Top three values of a derived quantity; ties are retained by default
HollywoodMovies |>
  mutate(score_difference = abs(RottenTomatoes - AudienceScore)) |>
  slice_max(order_by = score_difference, n = 3)
```


## Grouping and summaries

```{r}
#| eval: false
# Summarise one row per genre
HollywoodMovies |>
  group_by(Genre) |>
  summarise(mean_worldwide = mean(WorldwideBO, na.rm = TRUE), .groups = "drop")

# Summarise by genre and year; n() counts rows in each group
HollywoodMovies |>
  group_by(Genre, Year) |>
  summarise(
    mean_worldwide = mean(WorldwideBO, na.rm = TRUE),
    movie_count = n(),
    .groups = "drop"
  )
```


## Example: top 5 most profitable genres

```{r}
#| eval: false
HollywoodMovies |>
  filter(!is.na(Profitability)) |>
  group_by(Genre) |>
  summarise(
    mean_profit = mean(Profitability),
    movie_count = n(),
    .groups = "drop"
  ) |>
  slice_max(order_by = mean_profit, n = 5)
```

- Nothing is stored; no change is made in the data frame.


## Creating new variables

```{r}
#| eval: false
# Create a derived column and select the five largest values
HollywoodMovies |>
  mutate(score_gap = abs(RottenTomatoes - AudienceScore)) |>
  slice_max(order_by = score_gap, n = 5) |>
  select(Movie, RottenTomatoes, AudienceScore, score_gap)

# Save the updated data frame using explicit assignment
movies_with_gap <- HollywoodMovies |>
  mutate(score_gap = abs(RottenTomatoes - AudienceScore))

# Keep only selected columns while creating a derived column
HollywoodMovies |>
  transmute(Movie, score_gap = abs(RottenTomatoes - AudienceScore))

# Rename existing columns
HollywoodMovies |>
  rename(open_weekend_gross = OpenBO, theaters_at_open = TheatersOpen)
```



## Tibble

A **tibble** is a modern version of an R data frame. It prints a compact preview and preserves familiar data-frame behavior while integrating with tidyverse functions.

```{r}
#| eval: false
tbl_movie <- tibble::as_tibble(HollywoodMovies)
class(tbl_movie)
```

## Warm-up solutions

```{r}
#| eval: false
# Count movies by genre
HollywoodMovies |>
  count(Genre, name = "movie_count", sort = TRUE)

# Find the 10 largest absolute score differences
HollywoodMovies |>
  mutate(score_gap = abs(RottenTomatoes - AudienceScore)) |>
  slice_max(order_by = score_gap, n = 10) |>
  select(Movie, RottenTomatoes, AudienceScore, score_gap)
```

## Warm-up solutions (continued)

```{r}
#| eval: false
# Mean audience score by genre
HollywoodMovies |>
  group_by(Genre) |>
  summarise(
    mean_audience = mean(AudienceScore, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(desc(mean_audience))

# Five highest-grossing action movies
HollywoodMovies |>
  filter(Genre == "Action") |>
  slice_max(order_by = WorldwideBO, n = 5) |>
  select(Movie, Genre, WorldwideBO)
```

## Warm-up solutions (continued)

```{r}
#| eval: false
# Highest worldwide gross by genre (ties are retained)
HollywoodMovies |>
  group_by(Genre) |>
  slice_max(order_by = WorldwideBO, n = 1) |>
  ungroup() |>
  select(Movie, Genre, WorldwideBO)
```







