---
title: "DATA 352/551: Homework 2"
format:
  html:
    toc: false
---

Homework is not collected. The quiz based on this homework will be at the beginning of class, **Sep 16.**

## Question 1

Suppose a random variable $X$ can take values $\{1,2,3\}$ according to the probabilities $\{0.2,0.3,0.5\}$.

1. Simulate 10,000 observations from the distribution of $X$. Calculate the mean of the 10,000 simulated draws. Calculate the expectation of $X$ mathematically to compare.
2. Simulate another 10,000 observations from the distribution of $X$, with the condition that you are only allowed to make random draws *uniformly* from $(0,1)$. In R, you cannot use the `sample` function, but can only use `runif`.

<details>
<summary>Show solution</summary>

The mean should be very close to $0.2+0.3\times2+0.5\times3=2.3$.

```{r}
mysim <- sample(1:3, size = 10000,
                prob = c(0.2, 0.3, 0.5), replace = TRUE)
mean(mysim)
```

Suppose $U$ is a uniform random draw from $(0,1)$. To draw a value $x$ from the distribution of $X$:

- Set $x=1$ if $U<0.2$.
- Set $x=2$ if $0.2\le U<0.5$.
- Set $x=3$ if $0.5\le U$.

```{r}
sim_unif <- runif(10000)
sim_x <- sim_unif
sim_x[sim_unif < 0.2] <- 1
sim_x[sim_unif < 0.5 & sim_unif >= 0.2] <- 2
sim_x[sim_unif >= 0.5] <- 3
mean(sim_x)
```

</details>

## Question 2

There is a $2\%$ click rate for an online advertisement. The number of people that will click on the advertisement among the next $100$ people can be modeled by a binomial distribution.

1. What are the parameters of the binomial distribution?
2. Discuss the binomial model assumptions in the context of the problem.
3. Use your computer to calculate, among the next $100$ people: the probability that exactly two people click the advertisement; the probability that no one clicks the advertisement; and the probability that 5 or fewer people click the advertisement.

<details>
<summary>Show solution</summary>

We model it as a binomial distribution with parameters $n=100$ and $p=0.02$. There are two assumptions. (1) We assume that each person clicks the advertisement *independently* of each other; one person's action does not affect the action of the next person. (2) We assume that each person has the *same probability* of clicking the advertisement. This may be invalid if different populations have different click probabilities.

```{r}
dbinom(2, size = 100, prob = 0.02)  # P(exactly two)
dbinom(0, size = 100, prob = 0.02)  # P(no one)
pbinom(5, size = 100, prob = 0.02)  # P(five or fewer)
```

</details>

## Question 3

The number of times you get a 6 if you roll a fair, six-sided die for five times can be modeled by a binomial distribution.

1. What are the parameters of the binomial distribution?
2. Discuss the binomial model assumptions in the context of the problem. If the die were biased, can we still use a binomial distribution?
3. What is the probability that you get more than two 6's in five rolls?
4. Design a simulation to verify your answer in part 3, using the four steps in Lecture 1.

<details>
<summary>Show solution</summary>

We model it as a binomial distribution with parameters $n=5$ and $p=1/6$. We assume (1) each roll is independent of each other and (2) the probability of rolling a 6 is the same for each roll. We can still use a binomial model if the die is biased, as long as the probability of rolling a 6 stays the same.

```{r}
1 - pbinom(2, size = 5, prob = 1 / 6)  # P(more than two 6's)
```

For verifying the above probability:

- **One simulation/outcome:** Randomly sample five numbers from $\{1,2,\ldots,6\}$; for example, 2, 3, 2, 5, 6.
- **What to record:** The number of 6's in the five numbers.
- **Repeat many times:** Results might look like 1, 1, 2, 0, 3, ...
- **Summarize results:** Calculate the proportion of results greater than 2.

```{r}
number_of_sixes <- numeric(10000)
for (simulation in seq_len(10000)) {
  one_run <- sample(1:6, size = 5, replace = TRUE)
  number_of_sixes[simulation] <- sum(one_run == 6)
}
mean(number_of_sixes > 2)
```

</details>

## Question 4

A typesetter makes, on average, one mistake per 1000 word. Let $X$ be the number of mistakes that he makes on a page with $n$ words.

1. Suppose we model $X$ with a binomial distribution. What are the model parameters? Discuss the binomial model assumptions in the context of the problem.
2. With a binomial model, calculate the probabilities that: (i) he makes no mistake on a page with 100 words; (ii) he makes exactly 1 mistake on a page with 100 words; (iii) he makes 1 or more mistake on a page with 100 words; (iv) he makes no mistake on a page with 200 words.
3. Suppose we model $X$ with a Poisson distribution. What is the model parameter? Discuss the Poisson model assumptions in the context of the problem.
4. Re-calculate the probabilities in part 2 using a Poisson model.
5. Simulate 1000 random observations from the binomial model in part 1 and plot a histogram; simulate another 1000 random observations from the Poisson model in part 3 and also plot a histogram. Compare the two histograms.

<details>
<summary>Show solution</summary>

We can model $X$ with binomial $(n,p=0.001)$. We assume that whether he makes a mistake on one word is *independent* from whether he makes mistakes on other words, and that each word has *equal probability* that he makes a mistake.

```{r}
dbinom(0, size = 100, prob = 0.001)
dbinom(1, size = 100, prob = 0.001)
1 - dbinom(0, size = 100, prob = 0.001)
dbinom(0, size = 200, prob = 0.001)
```

We can also model $X$ with Poisson $(\lambda=0.001n)$. This is an approximation because the binomial sample space is $0,1,\ldots,n$, while the Poisson sample space is $0,1,\ldots,\infty$. We assume independent mistakes and that he makes mistakes at a *constant rate*. The probabilities are very close to the binomial model.

```{r}
dpois(0, lambda = 0.1)
dpois(1, lambda = 0.1)
1 - dpois(0, lambda = 0.1)
dpois(0, lambda = 0.2)
```

```{r}
par(mfrow = c(1, 2))
plot(as.factor(rbinom(1000, size = 100, prob = 0.001)),
     col = "lightblue", main = "binomial")
plot(as.factor(rpois(1000, lambda = 0.1)),
     col = "purple", main = "Poisson")
```

It is hard to estimate tail probabilities like $P(X=3,4,\ldots,100)$ using simulation because the probabilities are very small.

</details>

## Question 5

Do Question 18 on page 199 in *Introduction to Probability*.

> **18.** A baker blends 600 raisins and 400 chocolate chips into a dough mix and, from this, makes 500 cookies.
>
> 1. Find the probability that a randomly picked cookie will have no raisins.
> 2. Find the probability that a randomly picked cookie will have exactly two chocolate chips.
> 3. Find the probability that a randomly chosen cookie will have at least two bits (raisins or chips) in it.

<details>
<summary>Show solution</summary>

The number of raisins in a randomly picked cookie, $X$, can be modeled using a Poisson distribution with $\lambda=600/500$. We can also model $X$ with a binomial distribution with parameters $n=600$, $p=1/500$. The probabilities are very close, but the binomial model is more exact because it has the correct range $X\in\{0,\ldots,600\}$.

```{r}
dpois(0, lambda = 6 / 5)
dbinom(0, size = 600, prob = 1 / 500)
```

For part (b), use a Poisson model with $\lambda=400/500$:

```{r}
dpois(2, lambda = 4 / 5)
```

For part (c), use a Poisson model with $\lambda=1000/500$:

```{r}
1 - ppois(1, lambda = 2)
```

</details>

## Question 6

Under the setup of Question 26 on page 201 in *Introduction to Probability*, the number of hits in each of the 576 small areas can be modeled by a Poisson distribution. What is the parameter? Calculate the probabilities that a small area would have 0, 1, 2, 3, 4, and 5 or more hits under the Poisson model, and compare the probabilities to the actual proportions.

> **26.** Feller discusses the statistics of flying bomb hits in an area in the south of London during the Second World War. The area in question was divided into $24\times24=576$ small areas. The total number of hits was 537. There were 229 squares with 0 hits, 211 with 1 hit, 93 with 2 hits, 35 with 3 hits, 7 with 4 hits, and 1 with 5 or more. Assuming the hits were purely random, use the Poisson approximation to find the probability that a particular square would have exactly $k$ hits. Compute the expected number of squares that would have 0, 1, 2, 3, 4, and 5 or more hits and compare this with the observed results.

<details>
<summary>Show solution</summary>

The number of hits can be modeled by a Poisson distribution with $\lambda=537/576$. Calculation below shows that the Poisson model works very well.

```{r}
probability <- dpois(0:4, lambda = 537 / 576)
probability <- c(probability, 1 - ppois(4, lambda = 537 / 576))

expected <- 576 * probability
actual <- c(229, 211, 93, 35, 7, 1)

comparison <- cbind(expected, actual)
rownames(comparison) <- c(
  "0 hit", "1 hit", "2 hits", "3 hits", "4 hits", "5 or more hits"
)
round(comparison, 1)
```

</details>

## Question 7

Suppose that the probabilities of having a male or a female child are both 0.5, and you do not have to consider cases like twins. Calculate the following probabilities using appropriate models.

1. If you plan to have three children, what is the probability that you have exactly one daughter? What is the probability that you have at least one daughter?
2. What is the probability that your first daughter is your third child? (Hint: imagine the scenario of keeping having children until you have a daughter, and model the number of children when the first daughter “occurs.”)
3. What is the probability that your second daughter is your third child?
4. How many children do you need to have to guarantee with more than 95% chance that you will have at least a daughter?

<details>
<summary>Show solution</summary>

For parts (a), (b), and (c):

```{r}
dbinom(1, size = 3, prob = 0.5)       # a, exactly 1
1 - dbinom(0, size = 3, prob = 0.5)                   # a, at least 1

dgeom(2, prob = 0.5)                   # b
dnbinom(1, size = 2, prob = 0.5)       # c
```

For part (d), let $M$ be the number of children. The number of daughters, $X$, can be modeled by a binomial distribution with parameters $n=M$ and $p=0.5$. We want $P(X\ge1)=1-P(X=0)$ to be greater than $95\%$. We calculate this probability for different values of $M$; $M=5$ is the smallest number of children that satisfies the condition.

```r
1 - pbinom(0, size = 1:10, prob = 0.5)
```

</details>
