---
title: "Lecture 4: Common Continuous Distributions"
subtitle: "DATA 352/551: Modeling and Simulation"
engine: knitr
filters:
  - ../../../../_extensions/coatless/webr/webr.lua
format:
  revealjs:
    theme: default
    slide-number: true
    transition: fade
    toc: false
    self-contained: false
    smaller: false
    scrollable: true
    chalkboard:
         buttons: true
execute:
  echo: true
webr:
  autorun: false
editor_options: 
  chunk_output_type: inline
---

## Distribution name reference in R

![](distr){width=25%}


## Continuous random variables

Recall, a random variable $X$ is **continuous** if its sample space is continuous (and infinite). Can you think of some real life examples?

<br>

For discrete random variables, we calculate $P(X=x)$ for each possible value. We **cannot** do the same with continuous random variables. Why?

<br>

**Important property of a continuous random variable $X$**: $P(X=x) = 0$ for any value $x$. As a result:

- For example: $P(1 < X <3) = P(1 < X \le3) = P(1 \le X \le3)$.
- Only makes sense to calculate $P(X\in A)$ for some subset $A \subset \mathbb{R}$.



## Probability of continuous random variables

We can use [**histograms**](https://en.wikipedia.org/wiki/Histogram) to visualize *univariate*, continuous distributions (the discrete equivalent is called a [bar graph](https://en.wikipedia.org/wiki/Bar_chart)).

For the following continuous distributions, how would you calculate something like $P( a<X<b)$?

```{r}
#| code-fold: true
#| code-summary: "Show code"
#| out-width: 80%

par(mar=c(5.1,2,4.1,2),oma=c(0,0,2,0),mfrow=c(1,3))
hist(rnorm(n=1000,mean=162,sd=10),freq=F,
     ylab=NA,
     xlab='cm',
     main='Female height',
     yaxt='n',
     col='lightblue',
     #border='blue',lty='dashed',
     nclass=20,
     cex.lab=1.5,
     cex.axis=1.5,
     cex.main=1.5)

hist(rgamma(n=1000,shape = 2,scale = 28000),freq=F,
     xlab='dollars',
     ylab=NA,
     main='Annual income',
     #cex.main=2,
     yaxt='n',
     col='orange',
     #border='blue',lty='dashed',
     nclass=20,
     cex.lab=1.5,
     cex.axis=1.5,
     cex.main=1.5)

hist(faithful$waiting,xlab='min', main='Wait time b/t eruptions',probability=T, 
     col='lightgreen',nclass=20, yaxt='n', ylab=NA,
     cex.lab=1.5,
     cex.axis=1.5,
     cex.main=1.5)
```



## Density function

The probability mass function $f(x)=P(X=x)$ cannot be defined for continuous random variables. Instead, we use something called the **probability density function**, defined as a function $f(x)$ such that $P(X\in A)=\int_A f(x) dx$.

```{r}
#| code-fold: true
#| code-summary: "Show code"
#| out-width: 80%

par(mar=c(5.1,2,4.1,2),oma=c(0,0,2,0),mfrow=c(1,3))
hist(rnorm(n=1000,mean=162,sd=10),freq=F,
     ylab=NA,
     xlab='cm',
     main='Female height',
     yaxt='n',
     col='lightblue',
     #border='blue',lty='dashed',
     nclass=20,
     cex.lab=1.5,
     cex.axis=1.5,
     cex.main=1.5)
curve(dnorm(x,mean=162,sd=10),add=T,col='red',lwd=4)

hist(rgamma(n=1000,shape = 2,scale = 28000),freq=F,
     xlab='dollars',
     ylab=NA,
     main='Annual income',
     #cex.main=2,
     yaxt='n',
     col='orange',
     #border='blue',lty='dashed',
     nclass=20,
     cex.lab=1.5,
     cex.axis=1.5,
     cex.main=1.5)
curve(dgamma(x,shape = 2,scale=28000),add=T,col='red',lwd=4)

hist(faithful$waiting,xlab='min', main='Wait time b/t eruptions',probability=T, 
     col='lightgreen',nclass=20, yaxt='n', ylab=NA,
     cex.lab=1.5,
     cex.axis=1.5,
     cex.main=1.5)
dest = density(faithful$waiting,kernel = "gaussian",bw=2)
lines(dest, col = "red",lwd=4)
```


## Density estimation
A small digression: **density estimation** is a very big topic in statistics. We'll do a [simple demo](../code/L4_code.R) in R (R code: `density(data, kernel = "Gaussian", bw = 5)`).


<br>

For an *arbitrary* continuous density, calculating probabilities involves integration; R can handle **common continuous distributions** (i.e., their densities have certain functional forms, specified by given **parameters**); these are also used a lot in modeling.

## Uniform


## Uniform

**Scenario:** $X$ is a random draw from the interval $(a,b)$

**Parameter:** $a$ and $b$, end points of the range of $X$

**Mean:** $(a+b)/2$

**Probability:**

- **`dunif(2.3, min=0, max=3)`** does NOT give you probability!
- **`punif(2.3, min=0, max=3)`** gives $P(X\le2.3)$

**Simulation:**

- **`runif(1000, min=0, max=3)`**


## Exponential

## Exponential

**Scenario:** Exponential is often used to model *Poisson waiting time*: if the number of times an event occurs follows a Poisson distribution, then the waiting time between two events follows an exponential distribution.

>Question: if you stand in front of a window and spot taxis at a constant **rate** of 12.5 taxis per hour, how long would you **expect** to wait for the next taxi?

**Parameter:** $\lambda$ = rate of occurrence

**Mean:** $1/\lambda$ = expected or average waiting time

**Probability:**

- **`pexp(..., rate = ...)`**

**Simulation:**

- **`rexp(1000, rate = ...)`**



## Exercise

1. Suppose speeding cars occur at a constant rate. A policeman stands at a certain location and catches one speeding car per 10 minutes, on average. What is the chance that he has to wait for less than 5 minutes to catch the next speeding car? More than 15 minutes? Between 5 and 10 minutes?

<br><br>

2. Simulate from the above distribution; print out a few numbers and discuss what the numbers represent; plot a histogram to see its shape (note: the actual density function can be plotted by `curve(dexp(x, rate = ...))`).


## Beta

## Beta {.nostretch}

**Scenario:** Beta distribution has the range of $(0,1)$ and is often used to model percentages and probabilities, etc. (especially in Bayesian models as a prior distribution)

**Parameter:** $\alpha$ = shape 1; $\beta$ = shape 2

```{r}
#| code-fold: true
#| code-summary: "Show code"
#| fig-width: 12
#| fig-height: 3
#| out-width: 80%
#| fig-align: center

beta_shapes <- list(
     c(1, 1), c(1, 5), c(2, 2), c(2, 8), c(6, 4), c(0.2, 0.3)
)
beta_titles <- c(
     "alpha=1, beta=1", "alpha=1, beta=5", "alpha=2, beta=2",
     "alpha=2, beta=8", "alpha=6, beta=4", "alpha=0.2, beta=0.3"
)

par(mfrow = c(1, 6), mar = c(1, 0.5, 3, 0.5))
for (i in seq_along(beta_shapes)) {
     curve(
          dbeta(x, shape1 = beta_shapes[[i]][1], shape2 = beta_shapes[[i]][2]),
          from = 0.001, to = 0.999,
          col = "royalblue", lwd = 2,
          xlab = "", ylab = "", axes = FALSE,
          main = beta_titles[i]
     )
}
```

**Mean:** $\alpha / (\alpha + \beta)$

**Probability:**

- **`pbeta(..., shape1 = ..., shape2 = ...)`**

**Simulation:**

- **`rbeta(10000, shape1 = ..., shape2 = ...)`**



## Normal


## Normal {.nostretch}

**Scenario:** Normal distribution is typically used to model the sum, or the average, of "identical and independent" random variables (e.g., distribution of the sample mean), due to the *Central Limit Theorem*.

**Parameter:** $\mu$ = mean, $\sigma$ = standard deviation

```{r}
#| code-fold: true
#| code-summary: "Show code"
#| fig-width: 12
#| fig-height: 3
#| out-width: 80%
#| fig-align: center

normal_means <- c(0, 2.5, 162)
normal_sds <- c(1, 0.5, 10)
normal_titles <- c("mean=0,sd=1", "mean=2.5,sd=0.5", "mean=162,sd=10")

par(mfrow = c(1, 3), mar = c(4, 2, 3, 1))
for (i in seq_along(normal_means)) {
  sample_values <- rnorm(1000, mean = normal_means[i], sd = normal_sds[i])
  hist(sample_values, probability = TRUE,
       main = normal_titles[i], xlab = "", ylab = "",
       col = "lightblue", border = "blue", lty = "dashed",
       ylim = c(0, 1.1 * dnorm(normal_means[i],
                              mean = normal_means[i], sd = normal_sds[i])),
       xlim = normal_means[i] + c(-3, 3) * normal_sds[i])
  curve(dnorm(x, mean = normal_means[i], sd = normal_sds[i]),
        add = TRUE, col = "red", lwd = 2)
}
```

**Probability:**

- **`pnorm(..., mean = ..., sd = ...)`**

**Simulation:**

- **`rnorm(10000, mean = ..., sd = ...)`**

## Exercise

1. A community survey uses a Beta(6, 2) model for the proportion of residents who support a proposal. What is the probability that support exceeds 80%?

<br><br>

2. Exam scores are approximately normal with mean 80 and standard deviation 3. What proportion of students score between 75 and 85?

<br><br>

Notation for "distributed as": `% support ~ Beta(6, 2)`; `score ~ Normal(mean = 80, sd = 3)`


## The Central Limit Theorem

## Intuition: birthday example

If I pick a random student and ask their birth month, what number (between 1 and 12) would I get?

    A. A small number (like 1 or 2)

    B. A large number (like 11 or 12)

    C. Around 6

    D. No idea: any number is possible

Now, what if I ask a random sample of more than 1 students and calculate the **average** birth month? What would I get?



## Simulation

1. Suppose each of you is teaching a class with $n=25$ students. Simulate $25$ exam scores from uniform $(0,100)$ (try: plot a histogram of your simulated 25 exam scores; you should see a roughly uniform shape). Calculate the class **average** of your simulated class.

    Repeat for 10000 times, each time recording the class average. Plot a histogram of the 10000 simulated class averages. Estimate the mean and standard deviation (sd) of the class average?


2. Repeat 1, but simulating $25$ exam scores from normal with $mean=80$ and $sd=3$.

3. Repeat 1, but simulating $25$ exam scores from beta with $shape1=6$ and $shape2=2$.


> Your results should show that, regardless of your starting distribution (uniform, normal, or beta), the simulated class AVERAGES are very "well-behaved": the histogram of the 10000 simulated averages should look *normal* (symmetric bell-curve) in each case.


## The Central Limit Theorem

**Key result:** if you have a random sample of $n$ scores (these scores can be from ANY distribution!), then the **sample average** is random, and the randomness can be described by a **normal distribution** (for large $n$).

> **Why useful?** CLT enables us to make statistical inference based on the sample mean. For instance, you might make a *hypothesis* that the exam scores follow a beta distribution with shape1=6 and shape2=2.  To verify, you look at the average score of your 25 (randomly selected) students.  CLT tells you what this average **should** look like under your hypothesis.  If the observed average is "too extreme," you might decide to reject the hypothesis.

**Important question:** How large does $n$ have to be for CLT to work well? 

It depends on the shape of the population; if the population is very "not normal" (esp. long tail), $n$ needs to be very large to overcome the "un-normalness."


## Supplementary material

- (Recommended) [3Blue1Brown: But what is the Central Limit Theorem](https://www.youtube.com/watch?v=zeJD6dqJ5lo&t=532s)
- [Khan Academy: Central Limit Theorem lecture](https://www.youtube.com/watch?v=JNm3M9cqWyc&t=352s)
- [Interactive simulation](https://onlinestatbook.com/stat_sim/sampling_dist/)
