---
title: "Lecture 3: Implementing simulations"
subtitle: "DATA 352/551: Modeling and Simulation"
engine: knitr
filters:
  - ../../../../_extensions/coatless/webr/webr.lua
format:
  revealjs:
    theme: default
    slide-number: true
    transition: fade
    toc: false
    self-contained: false
    smaller: true
    scrollable: true
    chalkboard:
      buttons: true
execute:
  echo: true
webr:
  autorun: false
---

## Recommended workflow for simulations

1. Write a **dedicated function** for one simulation

> Input: parameters of the ONE simulation
> Output: a single result (i.e., "what do you want to save after ONE simulation?")        

2. Test the function with a few calls to make sure it works as expected

3. Initialize global simulation parameters

> How many simulations do you want to run?  How should the results be stored?
> Important: should always **pre-allocate** a vector (e.g. `rep(NA, 10000)`) to store results, rather than appending to a vector each time

4. Repeat the function many times
        
> Can use a `for` loop, or functional wrappers like `replicate()` (in R), or functions that can handle parallelization (i.e., use multiple cores on your computer to run simulations faster)



## Example: outcome of a fair die

What is the expected value of the outcome of a fair die?  Design and implement a simulation to find out. 

```{r}
# 1. ONE simulation function
sim_die <- function(n = 6) {
  return(sample(1:n,size=1))
}

# 2. Test the function
sim_die()

# 3. Initialize global simulation parameters
NSIM <- 10000
result <- rep(NA, NSIM)

# 4. Repeat the simulation NSIM times

#### Method 1: 'for' loop
for(i in 1:NSIM) {
  result[i] <- sim_die()
}
mean(result)


#### Method 2: Base R replicate()
result <- replicate(NSIM, sim_die())
mean(result)


#### Method 3: parallel processing (optional)
library(parallel)
cl <- makeCluster(2) # use 2 cores
clusterExport(cl, "sim_die")
result <- parSapply(cl, 1:NSIM, function(i) sim_die())
stopCluster(cl)
mean(result)
```


Note: Method 1 and Method 2 are computational similar (provided that the `result` vector is **pre-allocated**).  Method 3 is more advanced and can be significantly faster for very large simulations.

# R review

## Functions

An R function is a named piece of code that takes `inputs`, does some work, and returns an `output`.

```r
function_name <- function(input1, input2 = default_value) {
  # code that uses the inputs to generate result
  return(result)
}
```

- Give the function a descriptive name.
- Put the inputs in the parentheses, with defaults when useful.
- Write the calculation inside the braces.
- The last expression is returned automatically; `return()` is optional.
- To return more than one value, you can put the values in a `list`.

- Basic example 1:
```r
square_plus_one <- function(x) {
  return(x^2 + 1)
}

square_plus_one(3)
# 10
```


- Basic example 2:
```r
summarize_values <- function(first_value, second_value, third_value = 10) {
  result <- list(
    total = first_value + second_value + third_value,
    average = (first_value + second_value + third_value) / 3
  )
  return(result)
}

summarize_values(4, 8)
# $total
# [1] 22
#
# $average
# [1] 7.333333
```


- Exercise:

Write and test a function for each task:

1. `double_value(x)`: return twice the value of `x`.
2. `celsius_to_fahrenheit(celsius)`: convert a temperature using
   $F = 32 + 9C/5$.
3. `roll_die(n = 6)`: simulate one roll of a fair die with `n` sides.
4. `mean_of_two(x, y)`: return the average of two numbers.

<details>
<summary>Show solution</summary>

```{r}
double_value <- function(x) {
  result <- 2 * x
  return(result)
}

result <- double_value(4)
result

temperature <- function(celsius) {
  result <- 32 + 9 * celsius / 5
  return(result)
}
temperature(20)


roll_die <- function(n = 6) {
  result <- sample(1:n, size = 1)
  return(result)
}

roll_die()
roll_die(20)

mean_of_two <- function(x, y) {
  result <- (x + y) / 2
  return(result)
}

mean_of_two(3, 7)
```

</details>

## If statements

An `if` statement allows you to run a chunk of code only when a condition is true. It is often used together with loops.

Basic structure in R:

```r
if (condition) {
  # code to run when the condition is TRUE
}
```

You might need multiple `if` statements to cover more than one condition. You can also use an `else` statement.

Example: We roll a fair die. If the outcome is 6, we win $20; if the outcome is 4 or 5, we win $10; otherwise we lose $15. Is this a fair game? How much would you expect to win or lose? Write a function to play this game one time.

<details>
<summary>Show solution</summary>

```{r}
play_game <- function() {
  outcome <- sample(1:6, size = 1)

  if (outcome == 6) {
    result <- 20
  } else if (outcome %in% c(4, 5)) {
    result <- 10
  } else {
    result <- -15
  }

  return(result)
}

# One play
play_game()

# Approximate expected value by simulating many plays
NSIM <- 10000
result <- rep(NA, NSIM)
result <- replicate(NSIM, play_game())
mean(result)
```

For comparison, the expected value is

$$
E[X] = \frac{1}{6}(20) + \frac{2}{6}(10) + \frac{3}{6}(-15)
= \frac{20 + 20 - 45}{6}
= -\frac{5}{6}
\approx -0.83.
$$

So this is not a fair game: on average, you lose about $0.83 in the long run.

</details>


---

## Loops

Loops are used to repeat a chunk of code.

- `for` loop: repeat a fixed number of times
- `while` loop: repeat until a condition is met



## For loops

Basic example:
```r
for (i in 1:5) {
  print(i)
}
```

- A `for` loop has an index and a code chunk. For example, if the index is $i = 1, 2, 5$, then the code chunk runs with $i = 1$, then with $i = 2$, and then with $i = 5$. The index (`i` in the above example) does not need to be created separately.

Exercise: "fake coding"

1. Repeat a task 10 times.
2. Repeat a simulation 1000 times and store the result each time.
3. Suppose there is a large dataset of 1000 cats. Extract the 3rd, 6th, 9th, ... cats from the dataset; for each cat, run a function `weighCat()` and store the weight in a vector.
4. In a large dataset of cats, weigh and store the weights of the female cats.

Exercise: Use any language and loop structures to do the following.

1. Print a sequence of texts: “1 moose 4 legs”, “2 moose 8 legs”, ..., “30 moose 120 legs”.
2. Simulate the Martingale doubling system` ([Homework 1](../homework/homework_1.qmd)) 10 times. Start the betting amount at $1; if you win, you bet $1; if you lose, you double your bet. Assume a winning probability of $0.48$. For each of the 10 simulated games, print the game result (W or L), the current net winnings, and the current betting value.


---

## While loops

Basic example:
```r
x <- 1
while (x < 5) {
  print(x)
  x <- x + 1
}
```

A `while` loop has three ingredients:

- initialization
- condition
- code chunk that updates the condition

The code runs as long as the condition is met.  This is useful when the number of repetitions is not known in advance.

- It is important to properly update the objects involved in the while-loop condition.
- It is also recommended to set a safety upper bound so the loop cannot run forever.

```r
x <- 0
count <- 0
while (x < 5 && count < 1000) {
  x <- x + 1
  count <- count + 1
}
```

Exercise: "fake coding"

1. Play a game until you win $5$.
2. Play a game until you win $5$ or lose $100$.

Exercise: Use any language and loop structures to do the following.

1. Simulate the Martingale doubling system until you win $5$ or lose $100$. Start the betting amount at $1$; if you win, you bet $1$; if you lose, you double your bet. Assume a winning probability of $0.48$. For each simulated game, print the game result (W or L), the current net winnings, and the current betting value.
2. Write an R function to perform the above simulation.  Carefully design the inputs and outputs of the function.














