---
title: "STAT120 - Statistical Computing in R"
format:
  revealjs:
    theme: default
    slide-number: true
    transition: fade
    toc: false
    self-contained: false
    smaller: true
    scrollable: true
editor_options: 
  chunk_output_type: inline
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(fig.align='center',
                      fig.height=3,
                      fig.width=3,
                      out.width="50%",
                      echo=T)
```

## Introduction
- Wide variety, very flexible - can generate quite complicated graphs
  + http://www.htmlwidgets.org/showcase_leaflet.html
  + https://plot.ly/r/
- Many new packages and functions are being published (esp. on github)
- Currently, no "go-to" package, no unified approach
- We'll try a few simple examples to give you an idea, mostly using the <font color='red'>plotly</font> package

```{r,message=F}
#install.packages('plotly')
library(plotly)
library(ggplot2)
library(Lock5Data)
```


## Example 1: Histogram
- A histogram is given by default for plotting one variable
- Add <font color='blue'>type='histogram'</font> to get rid of the warning message
- Note the difference in syntax convention (e.g.: <font color='red'> ~ </font>varname)
```{r}
plot(mtcars$mpg)
```
</content>
