---
title: "Lecture 1: Introduction, numbers, and vectors"
format:
  revealjs:
    theme: default
    slide-number: true
    transition: fade
    toc: false
    self-contained: false
    smaller: true
    scrollable: true
editor_options: 
  chunk_output_type: inline
---


```{r setup, include = FALSE}
knitr::opts_chunk$set(fig.align='center',
                      fig.height=3,
                      fig.width=3,
                      out.width="50%",
                      echo=T)
```

# Getting started


## Posit cloud
Go to https://posit.cloud/ (you can log in with your Drew account).  We'll first use this online platform to demonstrate RStudio; later we'll download R and RStudio on your computer.

To start a new script, do **File - New File - R Script**.  <u>Do NOT choose "new project."</u>

Default layout of RStudio:

 + Upper left (R Script): this is a text file (.R format) that contains all the code you write
 + Lower left (Console): this is where your code is run (i.e., R will provide feedback and result here)
 + Upper right: contains information for the current R session
 + Lower right: contains information for files and packages; plots and help pages will also show up here
 

## Download R and RStudio

RStudio is an Integrated development environment (IDE) for R.  You need to download BOTH R and RStudio for RStudio to work.

- While you can use R without RStudio, the interface is not as nice
- When you open an R script (recall, this is a .R document), make sure it's opened in RStudio (to change the default: Right click - Get Info - Open with RStudio - Change All)
- Have a habit of saving R script often when you are coding
- When exit RStudio, you should **only save the R script** ("save workspace image?" - Don't Save)
- You can easily upload files to Posit Cloud, or download files from Posit Cloud

 
## Run R code
Running code directly in the console (1% of time):

- Fast and easy
- However: nothing is saved; can only run one line at a time

Instead (99% of time):

- Write code in R scripts
- Can save the R scripts to run or modify your code in the future 
- Can write many lines of code and run portions of it as needed


## Run R code

To run code in the console, you just need to **hit enter**; when you are writing code in the script, hitting enter will go to the next line but will not run the code.

To run code from R script:

- if you have **any highlighted code**
   + hit **command + enter** (mac) or **ctrl + enter** (windows)
   + (alternatively, you can click "Run")

- if you don't have highlighted code
   + hit **command/ctrl + enter**
   + R will **run the line of code** where your cursor is at (your cursor can be anywhere in the line)
   + R will also automatically move the cursor to the next line of code
   + This is a really handy way to run code line by line

## Try some basic math operations
To do $\sqrt{100}$ and $e^2$:
```{r,eval=F}
sqrt(100)
exp(2)
```

To do $\ln{(1000)}$ (log **base e** is the default, can change the base using `log(1000, base=10)`)
```{r, eval=F}
log(1000)
```

Exercise: calculate the following in R

1. $\frac{3\times 2 + 4^2}{5.5 \times 2}$
2. $\frac{e^{5.1}+23}{\frac{3}{4} \times 5}$
3. $\frac{\sqrt{12}}{\ln{(8-4.37)}}$


## R formatting basics
- Use \# to comment out text

- Space and empty lines do not matter in R

- R is case sensitive!
```{r,eval=F}
Sqrt(100)  # this will give you an error
```

- Use ';' to put multiple lines of code on the same line
```{r,eval=F}
2+4; 5*3
```

- Can separate one line of code into multiple lines, as long as it is obvious that the previous line has not ended
```{r,eval=F}
# works
3 + 2 - 7 *
  1000 + 10 - 100
# does not work
3 + 2 - 7
* 1000 + 10 - 100
```


## R packages
The ability to develop and use **R packages** is a big advantage of R over other commercial statistical software like SPSS or excel.

Before you can use a package, there are two steps:

(1) download and **install** the package: you only need to do this once per computer
    + How: click install under packages (recommended)
    + OR run `install.packages()`
(2) **load** the package: you need to do this each time you start a new R session
    + How: check the box under packages 
    + OR run `library()` (recommended)

```{r}
#install.packages('Lock5Data')
library('Lock5Data')
```

## Packages so far...
The **Lock5Data** package contains dataset from Lock, 
Lock, Lock, Lock, and Lock textbook *Statistics: Unlocking the Power of
Data*.  Other formats of the dataset can be downloaded from [http://www.lock5stat.com/datapage3e.html](http://www.lock5stat.com/datapage3e.html).

The **tidyverse** is a collection of very popular and powerful packages for data manipulation, visualization, formatting, etc. Details can be found at [https://www.tidyverse.org/packages/](https://www.tidyverse.org/packages/).  We'll talk more about these packages later. 


# R objects; numbers and vectors

## Creating an R object
- Examples of R object: a number, a vector, a data frame, a function, etc.

- To **assign** a number to an R object called "a":
```{r}
a=34
a = 34
a<-34
a <- 34
```
- These lines of code are equivalent - personal habit
- <u>Very useful</u>: you can print out an R object simply by running it
  + try: highlight "a" and run it


## Vector: a collection of values
- Example: `Cereal$Name` is a vector of 30 values
- When creating a vector, use `c( , )` for combined values
- Exercise: record information for a cereal that is not in the Cereal dataset and save it as an R object called **newCereal**


## R is case sensitive!
```{r, eval=F}
# these will give you errors
Plot(cereal$Carbs, cereal$Calories)
plot(cereal$Carbs, cereal$Calories)
plot(Cereal$carbs, Cereal$Calories)
```

- You will see these errors pretty often:
    + **Object `Reaction' not found**: check spelling; check where you thought you have created this object (did you actually create it by running the code?)
    + **Could not find function `Plot'**: check spelling; check if you have loaded relevant package
    + **'x' and 'y' lengths differ**: separately print out x and y to check




## Vector arithmetic
- When you apply some operation to a vector, R will perform the operation to *each element* in the vector 

```{r,eval=F}
2000 - Cereal$Calories
```

- Important note: the vector **Cereal$Calories** is not changed and the new values are not stored (verify this); if you want to store the new vector, you need to assign it to a new R object or override an existing R object
  
```{r,eval=F}
remainCalories = 2000 - Cereal$Calories
```

- When you do something like `vector1 ± vector2`, R will do the arithmetic operation *element by element*
  + if they have different lengths, R will give you a **warning**
  

## Basic vector functions
- Try these *functions*
```{r,eval=F}
mean(Cereal$Calories)
sum(Cereal$Calories)/length(Cereal$Calories)
range(Cereal$Calories)
min(Cereal$Calories)
sd(Cereal$Calories)
summary(Cereal$Calories)
```



## Character strings
- Put characters in double or single quotes
- R knows certain characters, like `col="red"`
```{r}
mycol="red"  #stores 'red' in an R object mycol
mycols=c('red','green','blue') #recall how to do vectors
```
- `mycol=red` $\rightarrow$ **Error: object 'red' not found**


## Logical vectors
- **TRUE** (or T) and **FALSE** (or F) are special logical characters when NOT in quotation
- **Numerically, R treats TRUE as 1 and FALSE as 0**
- Exercise: create a new vector, `cats`, that indicates if a student owns at least one cat or not
```{r, eval=F}
names <- c("Tom","Mary","Kate","James")
cats <- c(T,T,T,F)
as.numeric(cats)
```

- Q: what will **mean(cats)** return?

- Logical vectors are very useful for *subsetting* dataset.  For example, to get all the names of students who own cats:
```{r, eval=F}
names[cats]
```

## Logical vectors
- Logical vectors can be created using **logical operations**: $<, <=, >, >=, ==, !=$
```{r,eval=F}
Cereal$Calories < 100
Cereal$Company == 'K'
```

- **IMPORTANT: do not confuse `==` and `=`!**

- **which( )** will return the index of TRUE values
```{r,eval=F}
which(Cereal$Calories < 100)
which(Cereal$Company=='K')
```


# Vector modification and subsetting

## Append to a vector
- The easiest way is to use `c( , )` to combine vectors and/or a single element
```{r, eval=F}
# To append elements to a vector
names2=c("Aaron","Tom")
c(names,names2)
```

- The function **append** is more general. Run **?append** to get more information

- Nothing is stored!  Recall, to store, you have to assign the new vector to an R object.  We'll *override* the old vector to store the new vector:
```{r, eval=F}
names = c(names,names2)
cats = c(cats, F,F)
```

- Run **cbind(names,cats)** (*column bind*) to double check


## Subsetting: extract element(s) from a vector
- Use the **square bracket** for subsetting
```{r,eval=F}
#To get the 2nd person's name
names[2]

#To get the 2nd and 3rd persons' names
#name[2,3] does not work! Always use c( , ) for vectors!
names[c(2,3)]
```

Q: what will **names[c(2,2,2,3)]** return?


## Subsetting: extract element(s) from a vector
- Can also subset use a vector of TRUE and FALSE
```{r,eval=F}
names[cats]
```
- **R returns elements corresponding to TRUE**

- Alternatively, you can use the `which( )` function
```{r,eval=F}
Cereal$Name[Cereal$Protein>3]
Cereal$Name[which(Cereal$Protein>3)]
```



## Delete from a vector

Use the **minus sign** to delete the corresponding elements from a vector
```{r,eval=F}
names[-2]
names[-c(2,3)]
```


## Useful functions for creating vectors
```{r}
c(1:5)

# seq( ) is more general
seq(from=2,to=8,by=2)
seq(from=0,to=1,length.out=5)

rep(2,6)
```

```{r,eval=F}
#Useful for indexing
Cereal$Name[c(1:5)]
```

## Exercise: tallest buildings
- Go to https://en.wikipedia.org/wiki/List_of_tallest_buildings
- We will only look at the top 10 tallest buildings
  1. Create a vector that stores the Height (m) of the top 10 tallest buildings
  1. Use R to verify that the vector you created has a length of 10 
  1. Calculate the average Height in meters
  1. Convert height from meters to ft and store it in a new vector
  1. Print the values of the new vector and check that these values are the same as Height (ft) given in Wikipedia
  1. Create corresponding vectors for the number of floors and year built for the top 10 tallest buildings
  1. Create a vector of $\{0,1\}$'s, where 0 means the building was built in 2015 or before, and 1 means otherwise
  1. Generate a scatterplot of Floors by Height (m), set the color to "col=year_ind+2," where "year_ind" is the indicator vector created in the last step
  
## 
```{r}
#| code-fold: true
#| code-summary: "Show code"
#| fig-height: 4
#| fig-width: 4
#| out-width: "70%"
height_m=c(828, 632, 601, 599, 554.5, 541.3, 530, 530, 528, 508)
height_f=height_m * 3.28
floors=c(163, 128, 120, 115, 123, 104, 111, 98, 108, 101)
year=c(2010, 2015, 2012, 2017, 2016, 2014, 2016, 2018, 2018, 2004)
year.ind=as.numeric(year>2015)
plot(height_m,floors,col=year.ind+2,main='top 10 tallest buildings in the world',
     cex.lab=1.2,cex.axis=1.2,cex.main=1.2,pch=16)
legend('topleft',
       col=c(2,3),
       pch=c(16,16),
       legend=c('2015 and prior', '2016 and later'),cex=1.2)
```







