---
title: "Chapter 5: Bivariate Relationships"
subtitle: "STAT117: Introduction to Statistics"
format:
  revealjs:
    width: 1050
    height: 1050
    margin: 0
    theme: default
    slide-number: true
    transition: fade
    toc: false
    self-contained: false
    smaller: false
    scrollable: true
    chalkboard:
      buttons: true
    template-partials:
      - title-slide.html
editor_options:
  chunk_output_type: inline
---


## Introduction


Among runners ([StatChat discussion](https://youtu.be/xWy2PzISjco){target="_blank"}), is there a relationship between finishing time and gender? Between finishing time and BMI? Between age categories and runner types (e.g., elite, intermediate, recreational)?

In this chapter, we look at TWO variables.  We want to ask: do they have an **association**? For different groups of people (e.g., by gender or by age bracket), does the distribution of one variable (e.g., finishing time) differ? If there is an association, how strong is it?

How you answer these questions vary largely based on the **measurement type** of the variables.  You should use appropriate **statistics** (e.g., correlation coefficients and percentages comparisons) and **visualizations** (e.g., scatterplots).

⚠️ [**Correlation does not imply causation.**](https://en.wikipedia.org/wiki/Correlation_does_not_imply_causation){target="_blank"}



# Bivariate relationship WITHOUT nominal variables


## Scatterplot 📊

A scatterplot is a **bivariate** graph of two numeric variables. 

Example: In the `NELS` dataset, (`achrdg12`, `achmat12`) values can be considered in pairs:

- Student 1: (52.56, 59.69)
- Student 2: (53.12, 55.36)
- Student 3: (54.79, 56.73)
- $\ldots$

The scatterplot (next page) shows 500 dots, each corresponding to one student in the dataset. The highlighted dot represents Student 1.

## NELS achievement scatterplot

```{r}
#| echo: true
#| code-fold: true
#| code-summary: "R code (optional)"
#| fig-width: 6
#| fig-height: 6
#| out-width: "100%"
nels <- haven::read_sav("../../../../dataset/NELS.sav")

plot(
  nels$achrdg12,
  nels$achmat12,
  xlab = "Reading achievement in 12th grade (achrdg12)",
  ylab = "Math achievement in 12th grade (achmat12)",
  pch = 16,
  col = "gray70",
  asp = 1,
  cex = 1.5
)

abline(v = 52.56, lty = 2, col = "steelblue")
abline(h = 59.69, lty = 2, col = "steelblue")
points(52.56, 59.69, pch = 21, cex = 1.5, bg = "red", col = "black")
text(52.56, 59.69, labels = "Student 1", pos = 4, col = "red", cex=2)
```

## Other Scatterplot Examples

Here are more examples of scatterplots (see [StatChat2](https://youtu.be/F24R1WWqiKM){target="_blank"})

- [Best dogs](https://informationisbeautiful.net/visualizations/best-in-show-whats-the-top-data-dog/){target="_blank"}
- [Ranking of Large Language Models (LLMs)](https://informationisbeautiful.net/visualizations/the-rise-of-generative-ai-large-language-models-llms-like-chatgpt/){target="_blank"}
- [Is Sushi ‘Healthy’? What About Granola? Where Americans and Nutritionists Disagree](https://www.nytimes.com/interactive/2016/07/05/upshot/is-sushi-healthy-what-about-granola-where-americans-and-nutritionists-disagree.html?eafs_enabled=false){target="_blank"}


## Linear relationship

A scatterplot can visually display the linear relationship (positive or negative) between two numeric variables. The following graphs are based on the `AllCountries` dataset.

```{r}
#| echo: true
#| code-fold: true
#| code-summary: "R code (optional)"
#| fig-width: 12
#| fig-height: 7
#| out-width: "100%"
data("AllCountries", package = "Lock5Data")

par(mfrow = c(2, 3))

plot(
  AllCountries$CO2, AllCountries$Energy,
  xlab = "CO2", ylab = "Energy", pch = 16, cex = 0.7, col = "gray55",
  cex.lab = 1.5, cex.axis = 1.5
)
mtext("(a)", side = 3, line = 0.2, adj = 0, cex = 1.1, font = 2)

plot(
  AllCountries$Internet, AllCountries$LifeExpectancy,
  xlab = "Internet", ylab = "LifeExpectancy", pch = 16, cex = 0.7, col = "gray55",
  cex.lab = 1.5, cex.axis = 1.5
)
mtext("(b)", side = 3, line = 0.2, adj = 0, cex = 1.1, font = 2)

plot(
  AllCountries$Rural, AllCountries$Internet,
  xlab = "Rural", ylab = "Internet", pch = 16, cex = 0.7, col = "gray55",
  cex.lab = 1.5, cex.axis = 1.5
)
mtext("(c)", side = 3, line = 0.2, adj = 0, cex = 1.1, font = 2)

plot(
  AllCountries$Diabetes, AllCountries$Unemployment,
  xlab = "Diabetes", ylab = "Unemployment", pch = 16, cex = 0.7, col = "gray55",
  cex.lab = 1.5, cex.axis = 1.5
)
mtext("(d)", side = 3, line = 0.2, adj = 0, cex = 1.1, font = 2)

plot(
  AllCountries$GDP, AllCountries$BirthRate,
  xlab = "GDP", ylab = "BirthRate", pch = 16, cex = 0.7, col = "gray55",
  cex.lab = 1.5, cex.axis = 1.5
)
mtext("(e)", side = 3, line = 0.2, adj = 0, cex = 1.1, font = 2)
```


## Pearson Correlation Coefficient

The [Pearson correlation coefficient](https://en.wikipedia.org/wiki/Pearson_correlation_coefficient){target="_blank"}, denoted by $r$, is a statistic that measures the direction and strength of the linear relationship between two numeric variables.

- Direction: positive or negative (based on the sign of $r$)
- Strength: Cohen scale

| Absolute value of correlation | Strength |
| --- | --- |
| $0.50 \le |r| \le 1$ | Strong |
| $0.30 \le |r| < 0.50$ | Moderate |
| $0.10 \le |r| < 0.30$ | Weak |
| $|r| < 0.10$ | Negligible |


## Notes

- *Any* correlation coefficient is between $-1$ and $1$.
- $r$ can be strongly affected by outliers.
- $|r|$ (i.e. the magnitude of the correlation) is unchanged by a linear transformation of either variable, but the sign of $r$ changes if exactly one variable is multiplied by a negative constant.
- A steeper slope does not mean a stronger correlation.
- $r = 0$ only means no linear relationship, but a nonlinear relationship may still exist.


## Example

Download the [AllCountries](../../../../dataset/AllCountries.csv){target="_blank"} dataset and open it in Jamovi. Note that this is a CSV (comma-separated values) file, which is the preferred format for many statistical projects.

One downside is that descriptions of the variables are not included within the CSV file itself, but you can refer to the [AllCountries variable codebook](../../../../dataset/AllCountries_doc.html){target="_blank"}. It is a standard practice to provide data and its codebook separately.

Based on the `AllCountries` dataset, is there a relationship between carbon dioxide emissions (`CO2`) and energy use (`Energy`)?

<details>
<summary>Show solution</summary>

$r = 0.913$. There is a strong, positive linear relationship between carbon dioxide emissions and energy use: countries with higher CO2 emissions tend to use more energy, on average.
</details>

> Jamovi procedure: Analyses - Regression - Correlation Matrix - Put the two variables (`CO2` and `Energy` in this example) into the Variables box - (optional) check **Correlation matrix** under **Plot** to generate a scatterplot.




## Exercise

In the following bivariate relationship questions, support your answer with an appropriate statistic. If there is a linear relationship, describe its direction and strength using Cohen's scale. Interpret your answer in context.

Use the `AllCountries` dataset for these questions.

1. Is there a relationship between internet access (`Internet`) and life expectancy (`LifeExpectancy`)? 

2. Is there a relationship between the rural population percentage (`Rural`) and internet access (`Internet`)?

3. Is there a relationship between diabetes prevalence (`Diabetes`) and unemployment (`Unemployment`)? 



## Correlation Coefficient: Other Cases

Pearson $r$ is suitable for:

- Scale + scale
- Scale + dichotomous
- Dichotomous + dichotomous


**Spearman** correlation coefficient, denoted by $\rho$ ("rho"), is based on ranks and is suitable for:

- Ordinal + ordinal
- Ordinal + scale
- Ordinal + dichotomous

> Jamovi: Spearman $\rho$ is obtained using the same procedure as Pearson $r$, by selecting **Spearman** under the **Correlation Coefficients** option.

## Correlation Coefficients: Summary
When using a correlation coefficient, report:

- Direction
<br>

- Strength
<br>

- Interpretation in context
<br>

## Exercise (cont'd)

<strong>4.</strong> (`NELS`) Describe the direction and strength, using Cohen's scale, of the linear relationship between eighth-grade math achievement (`achmat08`) and teacher interest (`tcherint`).

<strong>5.</strong> (`NELS`) Is there an association between gender (`gender`; 0 = male, 1 = female) and eighth-grade math achievement (`achmat08`)?

<strong>6.</strong> (`United States Colleges`) Among schools in the `United States Colleges` dataset, is there an association between faculty size (`faculty`) and whether the school is a university (`univ`)?

<strong>7.</strong> (`United States Colleges`) Is there an association between affiliation (`public`; 0 = private, 1 = public) and institution type (`univ`; 0 = college, 1 = university)?

<strong>8.</strong> (`United States Colleges`) Is there an association between a school's competitive rank (`comp`) and its student-faculty ratio (`facratio`)?

<strong>9.</strong> (`United States Colleges`) Is there an association between a school's competitive rank (`comp`) and its faculty salary rank (`facsal`)?

<strong>10.</strong> (`United States Colleges`) Is there an association between a school's competitive rank (`comp`) and affiliation (`public`)?


# Bivariate relationship WITH nominal variables

## To be added

# Appendix

## Choice of method summary
Bivariate relationships without any nominal variable:

- **Scale + scale:** Pearson $r$
- **Scale + ordinal:** Spearman $\rho$
- **Scale + dichotomous:** Pearson $r$ (the point-biserial correlation); compare the scale-variable distributions across groups
- **Ordinal + ordinal:** Spearman $\rho$; compare ordinal-response distributions across groups; contingency table
- **Ordinal + dichotomous:** Spearman $\rho$; compare ordinal-response distributions across groups; contingency table
- **Dichotomous + dichotomous:** Pearson $r$ (the phi coefficient); contingency table

Bivariate relationships with at least one **nominal** variable:

- **Nominal + scale/ordinal:** compare outcome distributions across nominal groups
- **Nominal + dichotomous/nominal:** compare category percentages using a contingency table


**Visualizing relationships:**

- Scatterplots: two numerical variables
- Side-by-side boxplots: one numerical variable and one categorical variable
- Clustered bar graphs: two categorical variables
- Contingency tables: two categorical variables