---
title: "STAT120 - Statistical Computing in R"
subtitle: "Homework 1"
---


Download [R](https://cran.r-project.org/) and [RStudio](https://posit.co/download/rstudio-desktop/) locally on your computer and install them. Speak to me as soon as possible if you have any questions.


## Part I

1. Create and store a vector of values starting at 1 and ending at 103, with the values increasing in increments of 3.

<details>
<summary>Show solution</summary>

```{r}
v1 <- seq(from=1,to=103,by=3)
```
</details>


2. Add 4 to each of the values in the vector from problem 1.  Store the new vector.

<details>
<summary>Show solution</summary>

```{r}
v2 <- v1+4
```
</details>


3. Remove the 8th value from the vector in problem 2 and store the new 
vector.

<details>
<summary>Show solution</summary>

```{r}
v3 <- v2[-8]
```
</details>


4. Add the values 50 and 39 to the end of the vector in problem 3 and 
store the new vector.

<details>
<summary>Show solution</summary>

```{r}
v4 <- c(v3,50,39)
```
</details>

5. Find the sum of the elements of the vector from problem 4. Note: You should get 2,023.

<details>
<summary>Show solution</summary>

```{r}
sum(v4)
```
</details>



6. Using an `R` function, find the length of the vector in 
problem 5.

<details>
<summary>Show solution</summary>

```{r}
length(v4)
```
</details>


7. Create a character vector of length 3, `color.list`, that stores three colors: `deepskyblue`, `hotpink`, and `seagreen` (in this order).

<details>
<summary>Show solution</summary>

```{r}
color.list <- c('deepskyblue','hotpink','seagreen')
```
</details>


8. Create a numeric vector of length 100, `color.index`, that looks like $(1,2,1,2,1,2,\ldots)$. Hint: use the `rep()` function to repeat the vector $(1,2)$ 50 times.

<details>
<summary>Show solution</summary>

```{r}
color.index <- rep(c(1,2),50)
```
</details>


9. Use the vector from problem 8 to *index* the color vector in problem 7 and create a new vector that looks like ('deepskyblue', 'hotpink', 'deepskyblue', 'hotpink', ...)

<details>
<summary>Show solution</summary>

```{r}
mycol <- color.list[color.index]
mycol
```
</details>

10. Create a numeric sequence `x` that goes from 1 to 100; create another numeric sequence `y` that repeats the value 8 for 100 times. Use `R` to check that both `x` and `y` have lengths of 100.

<details>
<summary>Show solution</summary>

```{r}
x <- c(1:100)
y <- rep(8,100)
length(x)
length(y)
```
</details>

11. Use the vectors in problem 10 to perform the following calculation:
$${1\times8 + 2\times8 + 3\times8 + \ldots + 100 \times 8}.$$  You should get $40400$.

<details>
<summary>Show solution</summary>

```{r}
sum(x*y)
```
</details>


12. Create a scatterplot using the vectors in problem 10. Vector `x` should be on the x axis; vector `y` should be on the y axis. Set the color to the color vector you created in problem 9.

<details>
<summary>Show solution</summary>

```{r}
plot(x,y,col=mycol)
```
</details>


13. Do the scatterplot from problem 12 again, but add these arguments: set `xlab` and `ylab` to `NA`, set `type` to `"h"`, and set `ylim` to `c(5, 9)`. Pay attention to what each argument does.

<details>
<summary>Show solution</summary>

```{r}
plot(x,y,col=mycol,xlab=NA,ylab=NA,type='h',ylim=c(5,9))
```
</details>




# Part II

In this part, we'll use the built-in dataset `mtcars` to practice subsetting. Run the dataset to print it out and run `?mtcars` to bring up the help page, so that you know what the variables are.

As you see, the dataset contains 11 variables (`mpg`, `cyl`, `disp`, etc.), each of which is a numeric vector. To get one of these vectors, such as `mpg`, you can use `mtcars$mpg`. Alternatively, run `attach(mtcars)` first and then use `mpg` directly. In other words, you can use either syntax below to calculate the average `mpg`. If you decide to attach the dataset, you only need to do it once at the beginning.

```{r,eval=F,echo=T}
mean(mtcars$mpg)
### OR ###
attach(mtcars)
mean(mpg)
```


1. Run the following code, which creates a new character vector, `model`, that stores the models of the cars in the dataset.

```{r,echo=T,eval=T}
mtcars$model <- row.names(mtcars)
```


2. Print out `model` and `am` (transmission type) of the first car.

<details>
<summary>Show solution</summary>

```{r}
attach(mtcars)
model[1]
am[1]
```
</details>



3. Print out `model` and `am` of the first, second, and 10th car.

<details>
<summary>Show solution</summary>

```{r}
model[c(1,2,10)]
am[c(1,2,10)]
```
</details>


4. Print out `am` and `mpg` of "Toyota Corona".

<details>
<summary>Show solution</summary>

```{r}
am[model=='Toyota Corona']
mpg[model=='Toyota Corona']
```
</details>


5. Print `model` for all automatic (`am`) cars.

<details>
<summary>Show solution</summary>

```{r}
model[am==0]
```
</details>


6. Print `model` for all cars whose weight (`wt`) is larger than 5.

<details>
<summary>Show solution</summary>

```{r}
model[wt>5]
```
</details>

7. Print `model` and `mpg` for all cars whose `mpg` is smaller than 16.

<details>
<summary>Show solution</summary>

```{r}
data.frame(model = model[mpg<16], mpg = mpg[mpg<16])
```
</details>
