---
title: "STAT120 - Statistical Computing in R"
subtitle: "Homework 4"
---

## Part I

The dataset used for this part is [Video_Games_Sales_as_at_22_Dec_2016.csv](../../../dataset/Video_Games_Sales_as_at_22_Dec_2016.csv){download="Video_Games_Sales_as_at_22_Dec_2016.csv"}. You will practicing practice [tidyverse](https://tidyverse.org/) functions for data manipulation.


```{r, echo=F, eval=T, message=F}
library(tidyverse)
games <- read.csv("../../../dataset/Video_Games_Sales_as_at_22_Dec_2016.csv")
```

1. Import the dataset into R and call it **games**. We have been using the following functions to examine basic information about the dataset. Write the pipe version of the same code (i.e., use the same functions, but use the pipe operator to call each function).

```{r, eval=F, echo=T}
head(games)
str(games)
```

<details>
<summary>Show solution</summary>

```{r}
games %>% head()
games %>% str()
```

</details>

Note: check the output provided by **str()**. If **User_Score** and **Year_of_Release** are factors, run the following code to convert them to numeric vectors.

```{r, echo=T, eval=T}
games$Year_of_Release <- as.numeric(as.character(games$Year_of_Release))
games$User_Score <- as.numeric(as.character(games$User_Score))
```

2. The following code calculates the average game sales (in millions of units) in North America (NA), European Union (EU), and Japan (JP) for the games developed by Nintendo. Write the pipe version of the same code (i.e., use the same functions, but use the pipe operator to call each function).

```{r, eval=F, echo=T}
colMeans(subset(games, Developer == "Nintendo")[, c(6,7,8)])
```

<details>
<summary>Show solution</summary>

```{r}
games %>%
  subset(Developer == "Nintendo") %>%
  '['(, c(6, 7, 8)) %>%
  colMeans()
```

</details>

3. Subset the dataset so that the only games you see are those that belong to the "Shooter" genre. Print the data for the first five shooter games.

<details>
<summary>Show solution</summary>

```{r}
games %>%
  filter(Genre == "Shooter") %>%
  slice(1:5)
```

</details>

4. Subset the dataset so that the only variables you see are the name of the game, the platform on which the game runs, the year of release, and the genre. Print the data for the first five games.

<details>
<summary>Show solution</summary>

```{r}
games %>%
  select(Name, Platform, Year_of_Release, Genre) %>%
  slice(1:5)
```

</details>

5. Print the data for the top ten games in terms of the total number of units sold worldwide.

<details>
<summary>Show solution</summary>

```{r}
games %>%
  arrange(desc(Global_Sales)) %>%
  slice(1:10)
```

</details>

6. Subset the dataset so that the only games you see are those that are played on N64 (Nintendo 64) and that had at least one million units sold in North America (**NA_Sales** > 2). Print the **Name**, **Genre**, and **NA_Sales** number for these games.

<details>
<summary>Show solution</summary>

```{r}
games %>%
  filter(Platform == "N64", NA_Sales > 2) %>%
  select(Name, Genre, NA_Sales)
```

</details>

7. Create a new variable that measures the total number of units sold in North America and the European Union combined. Sort the data in terms of this new variable (from most copies to fewest copies), and then print the data for the top ten most sold games in NA and EU. You can print a few variables of your choice instead of all of the variables.

<details>
<summary>Show solution</summary>

```{r}
games %>%
  mutate(NA_EU_Sales = NA_Sales + EU_Sales) %>%
  arrange(desc(NA_EU_Sales)) %>%
  slice(1:10) %>%
  select(Name, Platform, Genre, NA_EU_Sales)
```

</details>

8. Find information for the top 10 role-playing games in terms of user score. Note that some games have user score recorded as "tbd", and you need to disregard these games.

<details>
<summary>Show solution</summary>

```{r}
games %>%
  filter(Genre == "Role-Playing", User_Score != "tbd") %>%
  arrange(desc(User_Score)) %>%
  slice(1:10) %>%
  select(Name, Platform, Year_of_Release, Genre, User_Score)
```

</details>

9. What is the most sold game globally for each of the game genres? You can print a few variables of your choice instead of all of the variables.

<details>
<summary>Show solution</summary>

```{r}
games %>%
  group_by(Genre) %>%
  top_n(1, Global_Sales) %>%
  select(Name, Genre, Global_Sales)
```

</details>

10. Which are the top 3 game genres with the highest global sales number, on average?

<details>
<summary>Show solution</summary>

```{r}
games %>%
  group_by(Genre) %>%
  summarise(globalMean = mean(Global_Sales)) %>%
  arrange(desc(globalMean))
```

</details>

11. How many games are in each of the genres?

<details>
<summary>Show solution</summary>

```{r}
games %>% count(Genre)
```

</details>

12. What are the top 10 games with the largest difference (in absolute value) between critics score and user score? Note that you need to disregard games with NA values for either of the scores (hint: use **filter**), and you need to multiply the user score by 10 so that the two scores are compared on the same scale. You can print a few variables of your choice instead of all of the variables.

<details>
<summary>Show solution</summary>

```{r}
games %>%
  filter(!is.na(Critic_Score), !is.na(User_Score)) %>%
  mutate(User_Score_10 = 10 * User_Score) %>%
  mutate(diff = abs(Critic_Score - User_Score_10)) %>%
  arrange(desc(diff)) %>%
  slice(1:10) %>%
  select(Name, Genre, Platform, Critic_Score, User_Score_10)
```

</details>

13. Repeat 12, but disregard "obscure" games by limiting to games that have sold more than 3 (million) copies globally.

<details>
<summary>Show solution</summary>

```{r}
games %>%
  filter(!is.na(Critic_Score), !is.na(User_Score), Global_Sales >= 3) %>%
  mutate(User_Score_10 = 10 * User_Score) %>%
  mutate(diff = abs(Critic_Score - User_Score_10)) %>%
  arrange(desc(diff)) %>%
  slice(1:10) %>%
  select(Name, Genre, Platform, Critic_Score, User_Score_10)
```

</details>

14. What are the top 10 platforms in terms of number of games released? How many games were released on each of those 10 platforms? What are the average critic scores of the games on the 10 platforms (ignoring games with missing scores)? What was the median release year for games on each of the 10 platforms (ignoring games with missing release year)?

<details>
<summary>Show solution</summary>

```{r}
platforms <- games %>%
  group_by(Platform) %>%
  summarize(
    number_of_games = n(),
    average_critic_score = mean(Critic_Score, na.rm = TRUE),
    median_release_year = median(Year_of_Release, na.rm = TRUE)
  ) %>%
  top_n(10, number_of_games)
platforms
```

</details>

```{r, echo=F, eval=F}
library(ggplot2)
games %>%
  filter(Platform %in% platforms$Platform) %>%
  ggplot(aes(x = Year_of_Release, y = User_Score, color = Platform, group = Platform)) +
  geom_vline(xintercept = 2010, linetype = 'dashed', color = 'grey') +
  geom_jitter(alpha = 0.01) +
  geom_smooth(method = "lm") +
  facet_wrap(~ Platform) +
  theme_minimal()
```

