---
title: "STAT120 - Statistical Computing in R"
subtitle: "Homework 2"
---

## Part I

1. Suppose the following are the final high school GPAs of a group of
college students: 3.62, 3.84, 3.95, 4.00, 3.12, 2.86, 3.77, 2.53, 3.26, and 2.90.  Create a vector that stores these values in the order specified.  Choose a variable name that's descriptive but relatively short.

<details>
<summary>Show solution</summary>

```{r}
library(tidyverse)
gpa <- c(3.62, 3.84, 3.95, 4.00, 3.12, 2.86, 3.77, 2.53, 3.26, 2.90)
```
</details>


2. For the vector you created in problem 1, print out which value is greater than 3.5 (i.e., the output should be: 1, 2,3,4,7)
<details>
<summary>Show solution</summary>

```{r}
which(gpa>3.5)
```
</details>


3. Use `R` to check how many students are recorded in the data set, what the maximum GPA is, and which student has the maximum GPA.
<details>
<summary>Show solution</summary>

```{r}
length(gpa)
max(gpa)
which(gpa==max(gpa))
```
</details>

4. Find the average GPA for this group of students.
<details>
<summary>Show solution</summary>

```{r}
mean(gpa)
```
</details>


5. Suppose the following are the SAT quantitative scores for the same
group of students: 640, 680, 790, 730, 580, 560, 670, 450, 650, and 490, 
respectively.  Create another vector that stores these SAT scores in the order 
provided.  Double check that it has the right length.
<details>
<summary>Show solution</summary>

```{r}
sat_quant <- c(640, 680, 790, 730, 580, 560, 670, 450, 650, 490)
length(sat_quant)
```
</details>

6. Print out the SAT quantitative scores for the 2nd and 3rd students. 
<details>
<summary>Show solution</summary>

```{r}
sat_quant[c(2,3)]
```
</details>


7. Print out the SAT quantitative scores for the students whose GPA is above 3.5. Avoid hard-coding: do not manually type `1, 2, 3, 4, 7` anywhere in your code.
<details>
<summary>Show solution</summary>

```{r}
sat_quant[gpa>3.5]
```
</details>


8. Create a vector for gender with the following values (in this order): 1, 1, 1, 0, 0, 0, 0, 0, 1, 1.  Note that 1 represents "male" and 0 represents "female."  Double check that it has the right length.

<details>
<summary>Show solution</summary>

```{r}
gender <- c(rep(1, 3), rep(0, 5), rep(1, 2))
length(gender)
```
</details>

9. Find the average of the values in the gender vector you created in problem 8.  Note that this represents the proportion of males in the group.  Whenever you have 0/1 data (and only when this is true!), the mean is the proportion of 1's.
<details>
<summary>Show solution</summary>

```{r}
mean(gender)
```
</details>


10. The respective years of the 10 students are: Senior, Freshman, Sophomore, Freshman, Junior, Sophomore, Sophomore, Freshman, Freshman, Senior.  Create a *factor* vector that stores the year of students (hint: create a character vector first and convert it to factor).  Use R to double check that its type is factor.

<details>
<summary>Show solution</summary>

```{r}
year <- c("Senior", "Freshman", "Sophomore", "Freshman", "Junior", "Sophomore", "Sophomore", "Freshman", "Freshman", "Senior")
year <- as.factor(year)
class(year)
```
</details>


11. Use the **tapply** function to print out the average GPA of students by Year.  Your output should look like this.

<details>
<summary>Show solution</summary>

```{r,eval=T}
library(tidyverse)
gpa <- c(3.62, 3.84, 3.95, 4.00, 3.12, 2.86, 3.77, 2.53, 3.26, 2.90)
year <- c("Senior", "Freshman", "Sophomore", "Freshman", "Junior", "Sophomore", "Sophomore", "Freshman", "Freshman", "Senior")
year <- as.factor(year)
tapply(gpa,year,mean)
```
</details>


12. Store the vectors you created previously (`gpa`, `sat_quant`, `gender`, and `year`) in a data frame using the `data.frame()` function. Then, use the `group_by()` function in `tidyverse` to get the average GPA of students by year.

<details>
<summary>Show solution</summary>

```{r}
df.student <- data.frame(gpa,year,sat_quant,gender)
df.student %>% 
  group_by(year) %>%
  summarise(mean(gpa))
```
</details>





## Part II

1. The dataset used in this part is [salaries-by-college-type-clean.csv](../../../dataset/salaries-by-college-type-clean.csv). Download the data and import it into R. Print the names of the variables.

```{r}
df.sal <- read.csv('../../../dataset/salaries-by-college-type-clean.csv')
str(df.sal)
```

2. Print the *names* of the 3rd, 4th, and 10th school. 
```{r}
df.sal$School.Name[c(3,4,10)]
```

3. Print all the information of the 3rd, 4th, and 10th school.
```{r}
df.sal[c(3,4,10),]
```


4. Calculate the average **Starting.Median.Salary** of the first 20 schools.

```{r}
mean(df.sal$Starting.Median.Salary[1:20])
```


5. Print out all information of the school with the highest **Starting.Median.Salary**?  

```{r}
df.sal[which.max(df.sal$Starting.Median.Salary),]
df.sal[which.min(df.sal$Starting.Median.Salary),]
```


6. Print out the *names* of schools with a **Starting.Median.Salary** of 65,000 or greater.

```{r}
myindex <- which(df.sal$Starting.Median.Salary >= 65000)
df.sal$School.Name[myindex]
```


7. Print out the *names* of schools with a **Starting.Median.Salary** of 65,000 or greater OR with a **Mid.Career.Median.Salary** of 130,000 or greater (hint: to do something like Condition 1 OR Condition 2, the code is Condition 1 \textbf{|} Condition 2).

```{r}
myindex <- which(df.sal$Starting.Median.Salary >= 65000 | df.sal$Mid.Career.Median.Salary >= 130000)
df.sal$School.Name[myindex]
```


8. Print out the *names* of schools with a **Starting.Median.Salary** of 65,000 or greater AND with a **Mid.Career.Median.Salary** of 130,000 or greater.

```{r}
myindex <- which(df.sal$Starting.Median.Salary >= 65000 & df.sal$Mid.Career.Median.Salary >= 130000)
df.sal$School.Name[myindex]
```


9. Print out the *names* of schools with a **Starting.Median.Salary** between 55,000 and 60,000 (hint: you cannot write code like $4<x<5$ in R.  You need to separately write $x>4$ and $x<5$).

```{r}
myindex <- which(df.sal$Starting.Median.Salary >= 55000 & df.sal$Starting.Median.Salary <= 60000)
df.sal$School.Name[myindex]
```


10. Print out the *names* of 'Party' or 'Engineering' schools.

```{r}
df.sal$School.Name[df.sal$School.Type %in% c('Party','Engineering')]
```


11. Print out the average **Starting.Median.Salary** and the average **Mid.Career.Median.Salary** by different school types.  Sort the output by the average starting median salary. 

```{r}
df.sal %>%
  group_by(School.Type) %>%
  summarise(start=mean(Starting.Median.Salary), mid=mean(Mid.Career.Median.Salary)) %>%
  arrange(start)
```